---
title: "Code Book for Acitivity Data Cleaning Project"
author: "Baoshi Sun"
date: "Friday, April 24, 2015"
output:
  html_document:
    keep_md: yes
---

## Data Loading

Before loading data, we should check the existence of data files. If not file is found, downloading will be performed.

```{r}
## Load necessary packages
library(dplyr)

## Load the Raw Data, if not exist, try to download and unzip
fileURL <- "https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"
zipfn <- "getdata-projectfiles-UCI HAR Dataset.zip"
destdir <- "UCI HAR Dataset"
if (!file.exists(destdir)) {
    if(!file.exists(zipfn)) {
        ## Download the data file
        download.file(fileURL, destfile=zipfn)
    }
    ## Unzip the data file
    unzip(zipfn)
}
```

Once we got the data files, they should be loaded into date frame:

* Data Description  
We need to load the features and labels as the common data descriptions for both testing data and training data.  
* Testing Data  
consists of  subject data, x data and y data.   
* Training Data   
consists of  subject data, x data and y data.   

```{r}
## Read txt file
### Read the activity labels
raw.labels <- read.table(paste(destdir, "/activity_labels.txt", sep=""))
### Read the feature list
raw.features <- read.table(paste(destdir, "/features.txt", sep=""))

### Read the testing datasets
raw.test.subject <- read.table(paste(destdir, "/test/subject_test.txt", sep=""), col.names=c("Subject"))
raw.test.x <- read.table(paste(destdir, "/test/X_test.txt", sep=""))
raw.test.y <- read.table(paste(destdir, "/test/y_test.txt", sep=""), col.names=c("Activity"))
dim(raw.test.subject); dim(raw.test.x); dim(raw.test.y)

### Read the training datasets
raw.train.subject <- read.table(paste(destdir, "/train/subject_train.txt", sep=""), col.names=c("Subject"))
raw.train.x <- read.table(paste(destdir, "/train/X_train.txt", sep=""))
raw.train.y <- read.table(paste(destdir, "/train/y_train.txt", sep=""), col.names=c("Activity"))
dim(raw.train.subject); dim(raw.train.x); dim(raw.train.y)
```

## Merge the testing data and the training data:   

Since the dataset consists of 3 parts (x, y and subject), the testing data and the training data should be combined repesctively. Then, the training data are appended to the tail of the testing data.

```{r}
## 1. Data Merge
### Combine subject, y & x for the test set
raw.test <- cbind(raw.test.subject, raw.test.y, raw.test.x)
### Combine subject, y & x for the train set
raw.train <- cbind(raw.train.subject, raw.train.y, raw.train.x)
### Merges the training and the test sets to create one data set
raw.data <- rbind(raw.test, raw.train)
dim(raw.data)
##str(raw.data)
```

## 2. Extracts only the measurements on the mean and standard deviation for each measurement.  

Only variables with the feature name containing mean(), Mean(), std() or Std() are picked. In this case, some regular expression syntax is applied. In addition, we also prepared the corresponding feature name list (colnames), which will be useful in followingr steps.

```{r}
### Put the raw data into dplyr data table
activity <- tbl_df(raw.data)
### mean() and std() column list, need to add up 2 (Subject, Activity)
collist <- grep("[Mm]ean\\()|[Ss]td\\()", raw.features$V2)
colnames <- grep("[Mm]ean\\()|[Ss]td\\()", raw.features$V2, value=TRUE)
collist <- collist + 2
collist <- c(1, 2, collist)
colnames <- c("Subject", "Activity", colnames)
### Select columns
activity <- activity[, c(collist)]
```

## 3. Uses descriptive activity names to name the activities in the data set

Here we use a for-loop to replace the label of activities with the actual names. In order to manipulate data, all labels are converted to charater class.

```{r}
### Change the Activity to charater class
activity$Activity <- as.character(activity$Activity)
raw.labels$V1 <- as.character(raw.labels$V1)
raw.labels$V2 <- as.character(raw.labels$V2)
for( i in 1:nrow(raw.labels) ) {
    activity$Activity <- gsub(raw.labels[i,1], raw.labels[i,2], activity$Activity)
}
```

## 4. Appropriately labels the data set with descriptive variable names.  

```{r}
names(activity) <- colnames
```

## 5. From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject.   

Tidy data is saved into a text file named 'tidydata.txt', which has the header (feature names) but no row names (according to the instruction). To verify the result, we can reload the file and check it's content.

```{r}
tidy.data <- group_by(activity, Subject, Activity) %>%
    summarise_each(funs(mean))
names(tidy.data) <- colnames

### Save the tidy dataset
write.table(tidy.data, "./tidydata.txt", row.name=FALSE)

### Load and test 
tidytest <- read.table("./tidydata.txt", header=TRUE)
dim(tidytest); names(tidytest)
#head(tidytest)
```

## Data Dictionary

The tidy data contains 180 records, corresponding to 30 subjects' average activity data for 6 types of activities.  

There are 68 variables in the tidy data. The first two variables stands for Subject and Activity respectively. And the rest of 66 variables are all the average of mean values and standard deviation values for data collected from different sensors.  

Subject  [integer]    - Identifies the subject who performed the activity  

Activity [character]  - Name of activity, including:   
            --    LAYING             
            --    SITTING            
            --    STANDING           
            --    WALKING           
            --    WALKING_DOWNSTAIRS    
            --    WALKING_UPSTAIRS   
                
tBodyAcc.mean...X       [numeric]  
tBodyAcc.mean...Y       [numeric]  
tBodyAcc.mean...Z       [numeric]  
tBodyAcc.std...X        [numeric]         
tBodyAcc.std...Y        [numeric]  
tBodyAcc.std...Z        [numeric]         
tGravityAcc.mean...X    [numeric]         
tGravityAcc.mean...Y    [numeric]     
tGravityAcc.mean...Z    [numeric]      
tGravityAcc.std...X     [numeric]     
tGravityAcc.std...Y     [numeric]      
tGravityAcc.std...Z     [numeric]     
tBodyAccJerk.mean...X   [numeric]      
tBodyAccJerk.mean...Y   [numeric]     
tBodyAccJerk.mean...Z   [numeric]      
tBodyAccJerk.std...X    [numeric]     
tBodyAccJerk.std...Y    [numeric]      
tBodyAccJerk.std...Z    [numeric]     
tBodyGyro.mean...X      [numeric]      
tBodyGyro.mean...Y      [numeric]     
tBodyGyro.mean...Z      [numeric]      
tBodyGyro.std...X       [numeric]     
tBodyGyro.std...Y       [numeric]      
tBodyGyro.std...Z       [numeric]     
tBodyGyroJerk.mean...X  [numeric]      
tBodyGyroJerk.mean...Y  [numeric]     
tBodyGyroJerk.mean...Z  [numeric]      
tBodyGyroJerk.std...X   [numeric]     
tBodyGyroJerk.std...Y   [numeric]      
tBodyGyroJerk.std...Z   [numeric]     
tBodyAccMag.mean..      [numeric]      
tBodyAccMag.std..       [numeric]     
tGravityAccMag.mean..   [numeric]      
tGravityAccMag.std..    [numeric]     
tBodyAccJerkMag.mean..  [numeric]      
tBodyAccJerkMag.std..   [numeric]     
tBodyGyroMag.mean..     [numeric]      
tBodyGyroMag.std..      [numeric]     
tBodyGyroJerkMag.mean.. [numeric]      
tBodyGyroJerkMag.std..  [numeric]     
fBodyAcc.mean...X       [numeric]      
fBodyAcc.mean...Y       [numeric]     
fBodyAcc.mean...Z       [numeric]      
fBodyAcc.std...X        [numeric]     
fBodyAcc.std...Y        [numeric]      
fBodyAcc.std...Z        [numeric]     
fBodyAccJerk.mean...X   [numeric]      
fBodyAccJerk.mean...Y   [numeric]     
fBodyAccJerk.mean...Z   [numeric]      
fBodyAccJerk.std...X    [numeric]     
fBodyAccJerk.std...Y    [numeric]      
fBodyAccJerk.std...Z    [numeric]     
fBodyGyro.mean...X      [numeric]      
fBodyGyro.mean...Y      [numeric]     
fBodyGyro.mean...Z      [numeric]      
fBodyGyro.std...X       [numeric]     
fBodyGyro.std...Y       [numeric]      
fBodyGyro.std...Z       [numeric]     
fBodyAccMag.mean..      [numeric]      
fBodyAccMag.std..       [numeric]     
fBodyBodyAccJerkMag.mean..  [numeric]    
fBodyBodyAccJerkMag.std..   [numeric]  
fBodyBodyGyroMag.mean..     [numeric]  
fBodyBodyGyroMag.std..      [numeric]  
fBodyBodyGyroJerkMag.mean.. [numeric]  
fBodyBodyGyroJerkMag.std..  [numeric]  
